### 2024-09-24 starter_folder-main/paper/paper.qmd
add literature and update PDF

---
title: "Water Quality and Safety at Toronto Beaches"
subtitle: "A Data-Driven Investigation into Seasonal Trends and Public Health Implications"
author: 
  - Ziyuan Shen
thanks: "Code and data are available at: https://github.com/Serena-SHEN1011/Toronto-Beach-Data"
date: 22 September 2024
date-format: long
abstract: "This study examines the water quality and environmental conditions of Toronto beaches based on daily observations collected from May to September each year from 2010 to 2023. The analyses focused on identifying key environmental factors that contribute to water safety warnings, such as temperature, waterFowl presence, and wave action. The results of the study indicate a strong link between rainfall, water temperature fluctuations and the issuance of beach advisories. This study provides valuable insights for policy makers and public health officials to enhance safety measures and management of Toronto's beaches during the peak summer months."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(kableExtra)
library(here)
library(tibble)
```

```{r}
#| include: false
#| warning: false
#| message: false

# Load the clean data
cleaned_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

# Introduction
Toronto's beaches are important recreational destinations for residents and visitors, especially during the summer period. However, the safety and quality of these beaches are largely influenced by everyday environmental conditions such as wind speed, air and water temperatures, and the presence of wildlife. These factors not only affect beach availability, but also have significant public health implications. For example, a study in PLOS ONE (@young2023recreational) investigated microbial contamination in recreational waters. Understanding these environmental factors is critical for public health officials to issue appropriate recommendations and ensure that beachgoers are aware of potential risks.

Despite ongoing efforts to monitor and manage beach conditions, there has been limited comprehensive analysis of how these environmental variables interact and affect overall beach safety. For example, a study published in Springer explored the correlation between environmental factors and recreational water illnesses, pointing to the need for better predictive models and interventions (@saleem2023same). Similarly, research in MDPI found that rainfall, water turbidity and the presence of waterfowl were often associated with increased bacterial contamination of beach waters, which further emphasises the importance of detailed analysis of beach conditions (@sanchez2021region). However, the specific relationships between wind speed and temperature changes on Toronto beaches remain under-explored.

There are over 18,000 daily observations of beach conditions in this dataset.

In the following section, I chose six key environmental variables to examine in depth: date, beach name, wind speed, air temperature, water temperature, and bird populations. The analysis is based on observations of over 18,000 daily beach conditions from the Toronto Open Data Portal. By using these variables, we explore how environmental factors affect beach safety and availability. Specifically, we examined the relationship between wind speed and water quality, bird populations and water pollution risk, and how temperature variations affect the overall beach experience. This analysis will help to better plan and communicate the safety and attractiveness of Toronto's beaches.

\newpage

# Data {#sec-data}
## Data Overview {#sec1-data} 
To investigate water quality and beach conditions in Toronto, I found the ‘Toronto Beach Observations’ dataset from the Open Data Toronto portal through the library opendatatoronto (@opendatatoronto). This dataset is provided by the City of Toronto and includes daily observations of water quality and beach conditions at Toronto's public beaches. It records water temperature, wind speed, and other important information critical to public health and recreational safety. A potential limitation of this dataset is the uncertainty of water quality sampling locations, as water quality conditions can change due to factors such as local weather and nearby pollution sources. Therefore, using aggregated data for daily conditions may not fully reflect the specific conditions at each beach. The raw data contains more than 18,000 observations of daily beach conditions and includes a number of variables, including date, beach name, wind speed, avian population, turbidity, etc. All variables are explained and described in (Section A.[1](#appendix-a.1)) In my this report, I will focuses on 8 key variables: date, beach name, wind's speed, air temperature, wether rain or not, water temperature, waterfowl, water turbidity.

I used R(@citeR) for all statistical computing and visualizing and associated software packages tidyverse(@tidyverse) to extensively for data wrangling, visualization, and manipulation. I also used dplyr(@dplyr) or the efficient processing of data, including filtering, summarising and selecting variables. Additionally, I leveraged kableExtra (@kableExtra) package to improve the tables in R Markdown documents. Similar methods have been used to assess environmental impacts on public spaces, for example, a ScienceDirect(@saleem2022validation) study examined the impact of environmental stressors on the availability of urban green space. Our analyses explored how environmental factors affect the safety and availability of beaches over time, with a particular focus on relationships such as the effect of wind speed on water turbidity and the impact of bird populations on pollution risk. This exhaustive approach provides valuable insights for public health officials and beach managers, helping to enhance the decision-making process regarding beach proposals.

## Cleaned Data {#sec2-data} 
During the data cleaning process, I took several important steps to ensure that the dataset was suitable for analysis. First, all rows containing missing values (NA) were removed. This decision was made to avoid any potential biases or inaccuracies that might result from incomplete data, as missing values could skew the results of the statistical analyses. By removing these rows, the dataset remained consistent and could be used for further analysis without introducing unnecessary complexity in dealing with missing data.

At the same time, several variables were removed because they were either redundant or irrelevant to the focus of this analysis. Wind direction was excluded because it has less impact on water conditions than wind speed. Similarly, rainfall was replaced by a simplified binary rainfall variable that indicates whether or not it rained on a given day. This binary approach was chosen to simplify the analysis without compromising the central idea that rainfall affects water quality. The role of waves was removed. This is because, although wave intensity affects beach conditions, its overall effect is not as important to the current discussion of weather, waterfowl, and water quality.

In addition, I have carefully selected nine key variables that were deemed most valuable to this analysis. These variables included date (the specific date on which observations were recorded) and beach name (identifying the different Toronto beaches on which environmental data were collected). Wind speed and air temperature were listed as key weather-related variables, as these factors have a significant impact on beach conditions and overall water quality. Water temperature was also considered critical because of its impact on both the beachgoer experience and biological activity in the water. Waterfowl population and turbidity were chosen because they are directly related to water quality. Waterfowl can cause pollution, while turbidity measures water clarity and is often used as an indicator of pollution. Finally, the presence or non-presence of rain was included to capture binary rainfall scenarios, as rainfall can affect water clarity and carry pollutants into the beach environment.

I cleaned and prepared a subset of the data as shown in @tbl-sum1 below. I showed the first and last 5 rows of the key variables in this table.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum1
#| fig-cap: "Displaying the first and last 5 rows of seven key variables"

# choose 8 variables
key_variables <- cleaned_data %>%
  select(dataCollectionDate, beachName, windSpeed, airTemp, rain, waterTemp, waterFowl, turbidity)

# Displaying the first and last 5 rows of key variables
sample_data <- bind_rows(head(key_variables, 5), tail(key_variables, 5))

# Generate table
kbl(sample_data, 
    col.names = c("Date", "Beach", "Wind (km/h)", "Air Temp (°C)", "Rain", "Water Temp (°C)", "Fowl Count", "Turbidity (NTU)"),
    ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center",
                font_size = 9) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%  
  column_spec(2, width = "8em") %>%  
  column_spec(3:4, width = "5em") %>%  
  column_spec(5, width = "2em") %>%  
  column_spec(6, width = "6em") %>%
  column_spec(7:8, width = "3em")

```

## Data Visualization {#sec3-data} 
  Air temperature is a key indicator of beach conditions. @fig-his1 visualizes the distribution of air temperature at Toronto beaches and the count of observations for different temperature ranges. The x-axis represents air temperature in degrees Celsius (°C), while the y-axis represents the count of observations. The distribution is approximately normal, centered around 20°C, indicating that most observations were recorded in this temperature range. There are fewer observations in the lower and higher temperature ranges, with temperatures ranging from around 5°C to nearly 35°C.
  The bar heights reflect the number of observations within each 2°C bin, showing that the majority of temperatures fall between 15°C and 25°C, typical of summer conditions. The steelblue color used for the bars provides a clear contrast against the black outlines, making the data easily distinguishable. This histogram effectively summarizes the air temperature data, revealing that Toronto beaches experience mild to warm temperatures during the observation period.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-his1
#| fig-cap: "Air Temperature Distribution"

ggplot(cleaned_data, aes(x = airTemp)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "black") +
  labs(x = "Air Temperature (°C)", y = "Count") +
  theme_minimal()

```

  Wind speed affects both the comfort of beachgoers and water quality. @fig-bar1 illustrates the average wind speed at different Toronto beaches. The x-axis represents the average wind speed in kilometers per hour (km/h), while the y-axis lists the names of the beaches. The chart is horizontally oriented for clarity, with the wind speed values increasing from left to right.The lightblue bars represent the average wind speed at each beach, and the bars are arranged in descending order of wind speed. The chart highlights the variability in wind conditions across different beaches, suggesting that some beaches are more exposed to wind, potentially affecting people's experiences and water conditions.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-bar1
#| fig-cap: "Average Wind Speed by Beach"

ggplot(cleaned_data %>% group_by(beachName) %>% summarize(avg_wind = mean(windSpeed, na.rm = TRUE)), 
       aes(x = reorder(beachName, avg_wind), y = avg_wind)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(x = "Beach", y = "Average Wind Speed (km/h)") +
  coord_flip() + 
  theme_minimal()

```

  Water temperature varies between beaches, which may influence swimmers' comfort and number of waterfowl in the water. @fig-bar2 shows the average water temperature at different Toronto beaches, with the x-axis representing the average water temperature in degrees Celsius (°C) and the y-axis listing the beach names. The chart is color-coded to distinguish each beach, and the bars are arranged in descending order of average water temperature.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-bar2
#| fig-cap: "Average water temperature by beach"
#| fig-subcap: ["Across different Toronto beaches"]

ggplot(cleaned_data %>% group_by(beachName) %>% summarize(avg_water_temp = mean(waterTemp, na.rm = TRUE)), 
       aes(x = reorder(beachName, avg_water_temp), y = avg_water_temp, fill = beachName)) +
  geom_col() +
  labs(x = "Beach", y = "Average Water Temperature (°C)") +
  coord_flip() +
  theme_minimal()  
```

  @fig-bar3 shows the total waterfowl count on rainy versus non-rainy days. The x-axis represents the presence or absence of rain, with three categories: "N/A" (unknown rain status), "No" (non-rainy days), and "Yes" (rainy days). The y-axis represents the total number of waterfowl observed. The chart indicates a noticeable difference in waterfowl presence based on weather conditions, with more waterfowl observed on non-rainy days compared to rainy ones. We can see that on non-rainy days (labeled "No"), the highest number of waterfowl were observed, with a count exceeding 60,000. On rainy days (labeled "Yes"), the total waterfowl count is significantly lower, around 40,000. The "N/A" category, representing days with unknown rain status, shows very few waterfowl, with a total count close to zero.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-bar3
#| fig-cap: "Total water fowl count on rainy vs non-rainy days"
#| results: hide  # This hides the console output but keeps the plot visible

# 2.# Bar plot: Total water fowl count on rainy vs non-rainy days
cleaned_data <- cleaned_data %>%
  mutate(rain = ifelse(is.na(rain), "Unknown", rain))  # Convert NA values to "Unknown"

# Checking the unique values of rain column
unique(cleaned_data$rain)

# Bar plot: Total water fowl count on rainy vs non-rainy days
fowl_rain_plot <- cleaned_data %>%
  group_by(rain) %>%
  summarize(total_water_fowl = sum(waterFowl, na.rm = TRUE)) %>%
  ggplot(aes(x = rain, y = total_water_fowl, fill = rain)) +
  geom_col() +
  labs(title = "Total Water Fowl Count on Rainy vs. Non-Rainy Days",
       subtitle = "Comparing the total number of water fowls observed on rainy vs. non-rainy days",
       x = "Rain (Yes/No)", y = "Total Water Fowl Count") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "gray")) +
  theme_minimal() +
  theme(legend.position = "none")

# Print plot and caption
print(fowl_rain_plot)

```
  @fig-scatter1 illustrates the relationship between wind speed (x-axis, measured in km/h) and turbidity (y-axis, measured in NTU). Each point represents an observation, with wind speed along the horizontal axis and turbidity along the vertical axis. This scatter plot shows that there is no strong linear relationship between wind speed and turbidity. However, lower wind speeds tend to be associated with a wide range of turbidity levels, whereas higher wind speeds appear to be less common and do not show higher turbidity levels. Most data points are clustered in areas with lower wind speeds (below 20 km/h) and turbidity levels below 50 NTU. Some outliers show higher turbidity levels above 150 NTU, especially at lower wind speeds. At wind speeds above 20 km/h, the data points become scattered, with very few data points over 40 km/h.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-scatter1
#| fig-cap: "Wind speed vs turbidity"
#| fig-subcap: ["Analyzing the relationship between wind speed and water turbidity"]

# Scatter plot: Wind speed vs turbidity
wind_turbidity_plot <- ggplot(cleaned_data, aes(x = windSpeed, y = turbidity)) +
  geom_point(alpha = 0.6, color = "orange") +
  labs(x = "Wind Speed (km/h)", y = "Turbidity (NTU)") +
  theme_minimal()

# Print plot and caption
print(wind_turbidity_plot)
```

  Rainfall significantly impacts water turbidity, as rain can introduce pollutants and sediments into the water. @fig-box1 shows the distribution of turbidity levels (measured in NTU) on rainy versus non-rainy days. The x-axis represents whether rain was present ("Yes"), absent ("No"), or unknown ("N/A"), while the y-axis represents turbidity levels. The box plot suggests that most observations, whether rainy or not, tend to have lower turbidity levels. Non-rainy days exhibit a slightly broader spread of turbidity levels, but the majority of observations have relatively low turbidity (close to 0 NTU), with several outliers reaching beyond 200 NTU. Turbidity levels were also mostly low in Rainy (Yes) and N/A(Unknown), but there were outliers over 150 NTU.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-box1
#| fig-cap: "Turbidity Levels on Rainy vs. Non-Rainy Days"

ggplot(cleaned_data, aes(x = rain, y = turbidity)) +
  geom_boxplot(aes(fill = rain)) +
  labs(x = "Rain (Yes/No)", y = "Turbidity (NTU)") +
  theme_minimal()
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-his2
#| fig-cap: "Distribution of water temperature"

# 5.Histogram: Distribution of water temperature
water_temp_hist <- ggplot(cleaned_data, aes(x = waterTemp)) +
  geom_histogram(binwidth = 20, fill = "blue", color = "black") +
  labs(x = "Water Temperature (°C)", y = "Count") +
  theme_minimal()

# Print plot and caption
print(water_temp_hist)

```

\newpage

# Appendix {#appendix-a}
## Variable description{#appendix-a.1}
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum2
#| fig-cap: "Description of variables in the dataset"

data <- data.frame(
  column = c("_id", "dataCollectionDate", "beachName", "windSpeed", 
             "windDirection", "airTemp", "rain", "rainAmount", "waterTemp", 
             "waterFowl", "waveAction", "waterClarity", "turbidity"),

  description = c("Unique row identifier for Open Data database", 
                  "Date observations were collected", 
                  "Name of beach where observations were collected", 
                  "Wind speed measured in km/h", 
                  "Wind directions - East, North-East, North, North-West, West, South-West, South, South-East", 
                  "Air temperature in degrees Celsius", 
                  "Presence of rain - Yes, No", 
                  "Amount of rainfall in mm in the last 24 hours", 
                  "Water temperature in degrees Celsius", 
                  "Count of water fowl seen", 
                  "Observations on wave action - high, low, moderate, none", 
                  "Visual state of water's clarity - free form text", 
                  "Turbidity of water measured in Nephelometric Turbidity Units (NTUs)")
)
data$description <- str_wrap(data$description, width = 25)

kable(data, 
      align = "l", 
      col.names = c("Column", "Description"),
      escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = F, 
                position = "center", 
                font_size = 10)
```

\newpage

# References

### 2024-09-24 starter_folder-main/paper/paper.qmd
update some changes

@@
 There are over 18,000 daily observations of beach conditions in this dataset.
-
 In the following section, I chose six key environmental variables to examine in depth: date, beach name, wind speed, air temperature, water temperature, and bird populations.
-The analysis is based on observations of over 18,000 daily beach conditions from the Toronto Open Data Portal.
 By using these variables, we explore how environmental factors affect beach safety and availability.
@@
 This analysis will help to better plan and communicate the safety and attractiveness of Toronto's beaches.
+Structure is inspired mainly by @rohan.
 
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-#| label: fig-his2
-#| fig-cap: "Distribution of water temperature"
+\newpage
 
-# 5.Histogram: Distribution of water temperature
-water_temp_hist <- ggplot(cleaned_data, aes(x = waterTemp)) +
-  geom_histogram(binwidth = 20, fill = "blue", color = "black") +
-  labs(x = "Water Temperature (°C)", y = "Count") +
-  theme_minimal()
-
-# Print plot and caption
-print(water_temp_hist)
-
-```
+# Discussion{#third}
+##
 

### 2024-09-24 starter_folder-main/paper/paper.tex
update some changes

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Water Quality and Safety at Toronto Beaches},
  pdfauthor={Ziyuan Shen},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Water Quality and Safety at Toronto Beaches\thanks{Code and data
are available at:
https://github.com/Serena-SHEN1011/Toronto-Beach-Data}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A Data-Driven Investigation into Seasonal Trends and Public
Health Implications}
\author{Ziyuan Shen}
\date{September 22, 2024}

\begin{document}
\maketitle
\begin{abstract}
This study examines the water quality and environmental conditions of
Toronto beaches based on daily observations collected from May to
September each year from 2010 to 2023. The analyses focused on
identifying key environmental factors that contribute to water safety
warnings, such as temperature, waterFowl presence, and wave action. The
results of the study indicate a strong link between rainfall, water
temperature fluctuations and the issuance of beach advisories. This
study provides valuable insights for policy makers and public health
officials to enhance safety measures and management of Toronto's beaches
during the peak summer months.
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\section{Introduction}\label{introduction}

Toronto's beaches are important recreational destinations for residents
and visitors, especially during the summer period. However, the safety
and quality of these beaches are largely influenced by everyday
environmental conditions such as wind speed, air and water temperatures,
and the presence of wildlife. These factors not only affect beach
availability, but also have significant public health implications. For
example, a study in PLOS ONE (Young et al. (2023)) investigated
microbial contamination in recreational waters. Understanding these
environmental factors is critical for public health officials to issue
appropriate recommendations and ensure that beachgoers are aware of
potential risks.

Despite ongoing efforts to monitor and manage beach conditions, there
has been limited comprehensive analysis of how these environmental
variables interact and affect overall beach safety. For example, a study
published in Springer explored the correlation between environmental
factors and recreational water illnesses, pointing to the need for
better predictive models and interventions (Saleem et al. (2023)).
Similarly, research in MDPI found that rainfall, water turbidity and the
presence of waterfowl were often associated with increased bacterial
contamination of beach waters, which further emphasises the importance
of detailed analysis of beach conditions (Sanchez et al. (2021)).
However, the specific relationships between wind speed and temperature
changes on Toronto beaches remain under-explored.

There are over 18,000 daily observations of beach conditions in this
dataset. In the following section, I chose six key environmental
variables to examine in depth: date, beach name, wind speed, air
temperature, water temperature, and bird populations. By using these
variables, we explore how environmental factors affect beach safety and
availability. Specifically, we examined the relationship between wind
speed and water quality, bird populations and water pollution risk, and
how temperature variations affect the overall beach experience. This
analysis will help to better plan and communicate the safety and
attractiveness of Toronto's beaches. Structure is inspired mainly by
Wickham et al. (2019a).

\newpage

\section{Data}\label{sec-data}

\subsection{Data Overview}\label{sec1-data}

To investigate water quality and beach conditions in Toronto, I found
the `Toronto Beach Observations' dataset from the Open Data Toronto
portal through the library opendatatoronto (Gelfand (2022)). This
dataset is provided by the City of Toronto and includes daily
observations of water quality and beach conditions at Toronto's public
beaches. It records water temperature, wind speed, and other important
information critical to public health and recreational safety. A
potential limitation of this dataset is the uncertainty of water quality
sampling locations, as water quality conditions can change due to
factors such as local weather and nearby pollution sources. Therefore,
using aggregated data for daily conditions may not fully reflect the
specific conditions at each beach. The raw data contains more than
18,000 observations of daily beach conditions and includes a number of
variables, including date, beach name, wind speed, avian population,
turbidity, etc. All variables are explained and described in (Section
A.\hyperref[appendix-a.1]{1}) In my this report, I will focuses on 8 key
variables: date, beach name, wind's speed, air temperature, wether rain
or not, water temperature, waterfowl, water turbidity.

I used R(R Core Team (2023)) for all statistical computing and
visualizing and associated software packages tidyverse(Wickham et al.
(2019b)) to extensively for data wrangling, visualization, and
manipulation. I also used dplyr(Wickham et al. (2023)) or the efficient
processing of data, including filtering, summarising and selecting
variables. Additionally, I leveraged kableExtra (Zhu (2024)) package to
improve the tables in R Markdown documents. Similar methods have been
used to assess environmental impacts on public spaces, for example, a
ScienceDirect(Saleem, Edge, and Schellhorn (2022)) study examined the
impact of environmental stressors on the availability of urban green
space. Our analyses explored how environmental factors affect the safety
and availability of beaches over time, with a particular focus on
relationships such as the effect of wind speed on water turbidity and
the impact of bird populations on pollution risk. This exhaustive
approach provides valuable insights for public health officials and
beach managers, helping to enhance the decision-making process regarding
beach proposals.

\subsection{Cleaned Data}\label{sec2-data}

During the data cleaning process, I took several important steps to
ensure that the dataset was suitable for analysis. First, all rows
containing missing values (NA) were removed. This decision was made to
avoid any potential biases or inaccuracies that might result from
incomplete data, as missing values could skew the results of the
statistical analyses. By removing these rows, the dataset remained
consistent and could be used for further analysis without introducing
unnecessary complexity in dealing with missing data.

At the same time, several variables were removed because they were
either redundant or irrelevant to the focus of this analysis. Wind
direction was excluded because it has less impact on water conditions
than wind speed. Similarly, rainfall was replaced by a simplified binary
rainfall variable that indicates whether or not it rained on a given
day. This binary approach was chosen to simplify the analysis without
compromising the central idea that rainfall affects water quality. The
role of waves was removed. This is because, although wave intensity
affects beach conditions, its overall effect is not as important to the
current discussion of weather, waterfowl, and water quality.

In addition, I have carefully selected nine key variables that were
deemed most valuable to this analysis. These variables included date
(the specific date on which observations were recorded) and beach name
(identifying the different Toronto beaches on which environmental data
were collected). Wind speed and air temperature were listed as key
weather-related variables, as these factors have a significant impact on
beach conditions and overall water quality. Water temperature was also
considered critical because of its impact on both the beachgoer
experience and biological activity in the water. Waterfowl population
and turbidity were chosen because they are directly related to water
quality. Waterfowl can cause pollution, while turbidity measures water
clarity and is often used as an indicator of pollution. Finally, the
presence or non-presence of rain was included to capture binary rainfall
scenarios, as rainfall can affect water clarity and carry pollutants
into the beach environment.

I cleaned and prepared a subset of the data as shown in
Table~\ref{tbl-sum1} below. I showed the first and last 5 rows of the
key variables in this table.

\begin{table}

\caption{\label{tbl-sum1}}

\centering{

\centering\begingroup\fontsize{9}{11}\selectfont

\begin{tabular}[t]{>{\raggedright\arraybackslash}p{6em}|>{\raggedright\arraybackslash}p{8em}|>{\raggedleft\arraybackslash}p{5em}|>{\raggedleft\arraybackslash}p{5em}|>{\raggedright\arraybackslash}p{2em}|>{\raggedleft\arraybackslash}p{6em}|>{\raggedleft\arraybackslash}p{3em}|>{\raggedleft\arraybackslash}p{3em}}
\hline
Date & Beach & Wind (km/h) & Air Temp (°C) & Rain & Water Temp (°C) & Fowl Count & Turbidity (NTU)\\
\hline
\textbf{2010-08-03} & Marie Curtis Park East Beach & 5 & 31 & Yes & 22.6 & 12 & 0.90\\
\hline
\textbf{2010-08-03} & Sunnyside Beach & 5 & 31 & Yes & 21.9 & 30 & 0.60\\
\hline
\textbf{2010-08-03} & Hanlan's Point Beach & 5 & 31 & Yes & 24.3 & 20 & 0.10\\
\hline
\textbf{2010-08-03} & Centre Island Beach & 5 & 31 & Yes & 21.3 & 30 & 0.20\\
\hline
\textbf{2010-08-03} & Cherry Beach & 10 & 30 & Yes & 21.3 & 10 & 1.30\\
\hline
\textbf{2019-06-28} & Marie Curtis Park East Beach & 8 & 18 & No & 13.0 & 3 & 0.88\\
\hline
\textbf{2019-06-28} & Sunnyside Beach & 8 & 18 & No & 12.0 & 13 & 1.01\\
\hline
\textbf{2019-07-30} & Marie Curtis Park East Beach & 9 & 18 & Yes & 11.0 & 12 & 2.38\\
\hline
\textbf{2019-07-30} & Sunnyside Beach & 9 & 18 & Yes & 14.0 & 50 & 2.88\\
\hline
\textbf{2023-06-21} & Sunnyside Beach & 30 & 20 & No & 19.0 & 8 & 0.91\\
\hline
\end{tabular}
\endgroup{}

Displaying the first and last 5 rows of seven key variables

}

\end{table}%

\subsection{Data Visualization}\label{sec3-data}

Air temperature is a key indicator of beach conditions.
Figure~\ref{fig-his1} visualizes the distribution of air temperature at
Toronto beaches and the count of observations for different temperature
ranges. The x-axis represents air temperature in degrees Celsius (°C),
while the y-axis represents the count of observations. The distribution
is approximately normal, centered around 20°C, indicating that most
observations were recorded in this temperature range. There are fewer
observations in the lower and higher temperature ranges, with
temperatures ranging from around 5°C to nearly 35°C. The bar heights
reflect the number of observations within each 2°C bin, showing that the
majority of temperatures fall between 15°C and 25°C, typical of summer
conditions. The steelblue color used for the bars provides a clear
contrast against the black outlines, making the data easily
distinguishable. This histogram effectively summarizes the air
temperature data, revealing that Toronto beaches experience mild to warm
temperatures during the observation period.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-his1-1.pdf}

}

\caption{\label{fig-his1}Air Temperature Distribution}

\end{figure}%

Wind speed affects both the comfort of beachgoers and water quality.
Figure~\ref{fig-bar1} illustrates the average wind speed at different
Toronto beaches. The x-axis represents the average wind speed in
kilometers per hour (km/h), while the y-axis lists the names of the
beaches. The chart is horizontally oriented for clarity, with the wind
speed values increasing from left to right.The lightblue bars represent
the average wind speed at each beach, and the bars are arranged in
descending order of wind speed. The chart highlights the variability in
wind conditions across different beaches, suggesting that some beaches
are more exposed to wind, potentially affecting people's experiences and
water conditions.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-bar1-1.pdf}

}

\caption{\label{fig-bar1}Average Wind Speed by Beach}

\end{figure}%

Water temperature varies between beaches, which may influence swimmers'
comfort and number of waterfowl in the water. Figure~\ref{fig-bar2}
shows the average water temperature at different Toronto beaches, with
the x-axis representing the average water temperature in degrees Celsius
(°C) and the y-axis listing the beach names. The chart is color-coded to
distinguish each beach, and the bars are arranged in descending order of
average water temperature.

\begin{figure}

\centering{

\centering{

\includegraphics{paper_files/figure-pdf/fig-bar2-1.pdf}

}

\subcaption{\label{fig-bar2}Across different Toronto beaches}

}

\caption{\label{fig-bar2}Average water temperature by beach}

\end{figure}%

Figure~\ref{fig-bar3} shows the total waterfowl count on rainy versus
non-rainy days. The x-axis represents the presence or absence of rain,
with three categories: ``N/A'' (unknown rain status), ``No'' (non-rainy
days), and ``Yes'' (rainy days). The y-axis represents the total number
of waterfowl observed. The chart indicates a noticeable difference in
waterfowl presence based on weather conditions, with more waterfowl
observed on non-rainy days compared to rainy ones. We can see that on
non-rainy days (labeled ``No''), the highest number of waterfowl were
observed, with a count exceeding 60,000. On rainy days (labeled
``Yes''), the total waterfowl count is significantly lower, around
40,000. The ``N/A'' category, representing days with unknown rain
status, shows very few waterfowl, with a total count close to zero.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-bar3-1.pdf}

}

\caption{\label{fig-bar3}Total water fowl count on rainy vs non-rainy
days}

\end{figure}%

Figure~\ref{fig-scatter1} illustrates the relationship between wind
speed (x-axis, measured in km/h) and turbidity (y-axis, measured in
NTU). Each point represents an observation, with wind speed along the
horizontal axis and turbidity along the vertical axis. This scatter plot
shows that there is no strong linear relationship between wind speed and
turbidity. However, lower wind speeds tend to be associated with a wide
range of turbidity levels, whereas higher wind speeds appear to be less
common and do not show higher turbidity levels. Most data points are
clustered in areas with lower wind speeds (below 20 km/h) and turbidity
levels below 50 NTU. Some outliers show higher turbidity levels above
150 NTU, especially at lower wind speeds. At wind speeds above 20 km/h,
the data points become scattered, with very few data points over 40
km/h.

\begin{figure}

\centering{

\centering{

\includegraphics{paper_files/figure-pdf/fig-scatter1-1.pdf}

}

\subcaption{\label{fig-scatter1}Analyzing the relationship between wind
speed and water turbidity}

}

\caption{\label{fig-scatter1}Wind speed vs turbidity}

\end{figure}%

Rainfall significantly impacts water turbidity, as rain can introduce
pollutants and sediments into the water. Figure~\ref{fig-box1} shows the
distribution of turbidity levels (measured in NTU) on rainy versus
non-rainy days. The x-axis represents whether rain was present
(``Yes''), absent (``No''), or unknown (``N/A''), while the y-axis
represents turbidity levels. The box plot suggests that most
observations, whether rainy or not, tend to have lower turbidity levels.
Non-rainy days exhibit a slightly broader spread of turbidity levels,
but the majority of observations have relatively low turbidity (close to
0 NTU), with several outliers reaching beyond 200 NTU. Turbidity levels
were also mostly low in Rainy (Yes) and N/A(Unknown), but there were
outliers over 150 NTU.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-box1-1.pdf}

}

\caption{\label{fig-box1}Turbidity Levels on Rainy vs.~Non-Rainy Days}

\end{figure}%

\newpage

\section{Discussion}\label{third}

\subsection{}\label{section}

\newpage

\section{Appendix}\label{appendix-a}

\subsection{Variable description}\label{appendix-a.1}

\begingroup\fontsize{10}{12}\selectfont

\begin{longtable}[t]{ll}

\caption{\label{tbl-sum2}}

\tabularnewline

\toprule
Column & Description\\
\midrule
\_id & Unique row identifier for
Open Data database\\
dataCollectionDate & Date observations were
collected\\
beachName & Name of beach where
observations were
collected\\
windSpeed & Wind speed measured in
km/h\\
windDirection & Wind directions - East,
North-East, North,
North-West, West,
South-West, South,
South-E\\
\addlinespace
airTemp & Air temperature in
degrees Celsius\\
rain & Presence of rain - Yes,
No\\
rainAmount & Amount of rainfall in mm
in the last 24 hours\\
waterTemp & Water temperature in
degrees Celsius\\
waterFowl & Count of water fowl seen\\
\addlinespace
waveAction & Observations on wave
action - high, low,
moderate, none\\
waterClarity & Visual state of water's
clarity - free form text\\
turbidity & Turbidity of water
measured in Nephelometric
Turbidity Units (NTUs)\\
\bottomrule

\end{longtable}

\endgroup{}

Description of variables in the dataset

\newpage

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-opendatatoronto}
Gelfand, Sharla. 2022. \emph{Opendatatoronto: Access the City of Toronto
Open Data Portal}.
\url{https://CRAN.R-project.org/package=opendatatoronto}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2023. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-saleem2022validation}
Saleem, Faizan, Thomas A Edge, and Herb E Schellhorn. 2022.
{``Validation of qPCR Method for Enterococci Quantification at Toronto
Beaches: Application for Rapid Recreational Water Monitoring.''}
\emph{Journal of Great Lakes Research} 48 (3): 707--16.

\bibitem[\citeproctext]{ref-saleem2023same}
Saleem, Faizan, Herb E Schellhorn, Albert Simhon, and Thomas A Edge.
2023. {``Same-Day Enterococcus qPCR Results of Recreational Water
Quality at Two Toronto Beaches Provide Added Public Health Protection
and Reduced Beach Days Lost.''} \emph{Canadian Journal of Public Health}
114 (4): 676--87.

\bibitem[\citeproctext]{ref-sanchez2021region}
Sanchez, Johanna, Jordan Tustin, Cole Heasley, Mahesh Patel, Jeremy
Kelly, Anthony Habjan, Ryan Waterhouse, and Ian Young. 2021.
{``Region-Specific Associations Between Environmental Factors and
Escherichia Coli in Freshwater Beaches in Toronto and Niagara Region,
Canada.''} \emph{International Journal of Environmental Research and
Public Health} 18 (23): 12841.

\bibitem[\citeproctext]{ref-rohan}
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019a.
{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\bibitem[\citeproctext]{ref-tidyverse}
---------, et al. 2019b. {``Welcome to the {tidyverse}.''} \emph{Journal
of Open Source Software} 4 (43): 1686.
\url{https://doi.org/10.21105/joss.01686}.

\bibitem[\citeproctext]{ref-dplyr}
Wickham, Hadley, Romain François, Lionel Henry, Kirill Müller, and Davis
Vaughan. 2023. \emph{Dplyr: A Grammar of Data Manipulation}.
\url{https://CRAN.R-project.org/package=dplyr}.

\bibitem[\citeproctext]{ref-young2023recreational}
Young, Ian, J Johanna Sanchez, Binyam Negussie Desta, Cole Heasley, and
Jordan Tustin. 2023. {``Recreational Water Exposures and Illness
Outcomes at a Freshwater Beach in Toronto, Canada: A Prospective Cohort
Pilot Study.''} \emph{PLoS One} 18 (6): e0286584.

\bibitem[\citeproctext]{ref-kableExtra}
Zhu, Hao. 2024. \emph{kableExtra: Construct Complex Table with 'Kable'
and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.

\end{CSLReferences}

\end{document}

### 2024-09-24 starter_folder-main/paper/paper.qmd
update some changes

@@
 
-# Discussion{#third}
-##
+# Discussion{#third-dis}
+## Data Collection and Limitations
+Observations in the dataset are obtained by city staff through a combination of both manual estimates and instrumental measurements.
+While the use of automatic tools such as thermometers and turbidimeters can provide more objective and consistent data, manual observations such as waterfowl counts and qualitative assessments of water clarity introduce an element of subjectivity.
+In addition, removing outliers and missing data entries during the clean-up process addresses errors that can occur during data collection, especially extreme or unreasonable values (e.g., water temperatures above 200°C).
+
+In summary, we measured real-world temperature, wind, rainfall, turbidity, and waterfowl activity using a variety of instruments and estimation techniques, resulting in quantifiable data entries in the dataset.
+These measurements form the basis for analysing environmental conditions at Toronto Beach, and transform natural phenomena into structured data that can be analysed to reveal trends, patterns and relationships between different variables.
 

### 2024-09-24 starter_folder-main/paper/paper.qmd
update

@@
 ---
-title: "Water Quality and Safety at Toronto Beaches"
-subtitle: "A Data-Driven Investigation into Seasonal Trends and Public Health Implications"
+title: "Environmental Factors Affecting Water Quality at Toronto Beaches"
+subtitle: "Investigation into Seasonal Trends"
 author: 
@@
 
-In summary, we measured real-world temperature, wind, rainfall, turbidity, and waterfowl activity using a variety of instruments and estimation techniques, resulting in quantifiable data entries in the dataset.
+In summary, we measured real-world temperature, speed of wind, rainfall, turbidity, and waterfowl activity using a variety of instruments and estimation techniques, resulting in quantifiable data entries in the dataset.
 These measurements form the basis for analysing environmental conditions at Toronto Beach, and transform natural phenomena into structured data that can be analysed to reveal trends, patterns and relationships between different variables.

### 2024-09-24 starter_folder-main/paper/paper.qmd
update

@@
 title: "Environmental Factors Affecting Water Quality at Toronto Beaches"
-subtitle: "Investigation into Seasonal Trends"
+subtitle: "Investigation into Seasonal Trends(2010-2023)"
 author: 
@@
 thanks: "Code and data are available at: https://github.com/Serena-SHEN1011/Toronto-Beach-Data"
-date: 22 September 2024
+date: 23 September 2024
 date-format: long

### 2024-09-27 starter_folder-main/paper/paper.qmd
remove some outliers

--- 2024-09-24 starter_folder-main/paper/paper.qmd
@@
 thanks: "Code and data are available at: https://github.com/Serena-SHEN1011/Toronto-Beach-Data"
-date: 22 September 2024
+date: 23 September 2024
 date-format: long
-abstract: "This study examines the water quality and environmental conditions of Toronto beaches based on daily observations collected from May to September each year from 2010 to 2023.
-The analyses focused on identifying key environmental factors that contribute to water safety warnings, such as temperature, waterFowl presence, and wave action.
-The results of the study indicate a strong link between rainfall, water temperature fluctuations and the issuance of beach advisories.
-This study provides valuable insights for policy makers and public health officials to enhance safety measures and management of Toronto's beaches during the peak summer months."
+abstract: "This paper investigates the environmental factors that influence water quality at Toronto beaches using summer data collected from 2010 to 2023.
+The research shows key variables such as rainfall, water temperature and waterfowl populations that significantly impact water quality safety warnings and public health.
+The study found that rainfall and water temperature fluctuations were strongly associated with beach warnings, emphasizing the necessity for improved safety measures.
+This research will help improve management strategies for Toronto's beaches and ensure a safer recreational experience for residents and visitors."
 format: pdf
@@
 There are over 18,000 daily observations of beach conditions in this dataset.
-In the following section, I chose six key environmental variables to examine in depth: date, beach name, wind speed, air temperature, water temperature, and bird populations.
+In the following section, I chose eight key environmental variables to examine in depth: date, beach name, wind speed, air temperature, water temperature, and waterfowl populations.
 By using these variables, we explore how environmental factors affect beach safety and availability.
@@
 To investigate water quality and beach conditions in Toronto, I found the ‘Toronto Beach Observations’ dataset from the Open Data Toronto portal through the library opendatatoronto (@opendatatoronto).
-This dataset is provided by the City of Toronto and includes daily observations of water quality and beach conditions at Toronto's public beaches.
-It records water temperature, wind speed, and other important information critical to public health and recreational safety.
-A potential limitation of this dataset is the uncertainty of water quality sampling locations, as water quality conditions can change due to factors such as local weather and nearby pollution sources.
-Therefore, using aggregated data for daily conditions may not fully reflect the specific conditions at each beach.
-The raw data contains more than 18,000 observations of daily beach conditions and includes a number of variables, including date, beach name, wind speed, avian population, turbidity, etc.
+The dataset contains more than 18,000 observations of daily beach conditions and includes a number of various environmental variables, including air temperature, water temperature, wind speed, and waterfowl population, which play a significant role in determining beach water safety.
+The analysis focuses on the key variables affecting water turbidity, which is a proxy for water pollution risk.
 All variables are explained and described in (Section A.[1](#appendix-a.1)) In my this report, I will focuses on 8 key variables: date, beach name, wind's speed, air temperature, wether rain or not, water temperature, waterfowl, water turbidity.
@@
 I also used dplyr(@dplyr) or the efficient processing of data, including filtering, summarising and selecting variables.
-Additionally, I leveraged kableExtra (@kableExtra) package to improve the tables in R Markdown documents.
+Additionally, I leveraged kableExtra (@kableExtra) package to improve the tables in R Markdown documents and used (@here) package to simplify file path management.
+It provides a more reliable and reproducible way to construct file paths by using the project root directory as a reference point.
 Similar methods have been used to assess environmental impacts on public spaces, for example, a ScienceDirect(@saleem2022validation) study examined the impact of environmental stressors on the availability of urban green space.
-Our analyses explored how environmental factors affect the safety and availability of beaches over time, with a particular focus on relationships such as the effect of wind speed on water turbidity and the impact of bird populations on pollution risk.
+My analyses explored how environmental factors affect the safety and availability of beaches over time, with a particular focus on relationships such as the effect of wind speed on water turbidity and the impact of bird populations on pollution risk.
 This exhaustive approach provides valuable insights for public health officials and beach managers, helping to enhance the decision-making process regarding beach proposals.
@@
   column_spec(7:8, width = "3em")
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: tbl-sum5
+#| fig-cap: ""
+
+# Calculate summary statistics for key variables
+summary_stats <- key_variables %>%
+  summarise(
+    WindSpeed_Mean = round(mean(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_SD = round(sd(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_Min = round(min(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_Median = round(median(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_Max = round(max(windSpeed, na.rm = TRUE), 1),
+
+    AirTemp_Mean = round(mean(airTemp, na.rm = TRUE), 1),
+    AirTemp_SD = round(sd(airTemp, na.rm = TRUE), 1),
+    AirTemp_Min = round(min(airTemp, na.rm = TRUE), 1),
+    AirTemp_Median = round(median(airTemp, na.rm = TRUE), 1),
+    AirTemp_Max = round(max(airTemp, na.rm = TRUE), 1),
+
+    WaterTemp_Mean = round(mean(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_SD = round(sd(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_Min = round(min(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_Median = round(median(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_Max = round(max(waterTemp, na.rm = TRUE), 1),
+
+    FowlCount_Mean = round(mean(waterFowl, na.rm = TRUE), 1),
+    FowlCount_SD = round(sd(waterFowl, na.rm = TRUE), 1),
+    FowlCount_Min = round(min(waterFowl, na.rm = TRUE), 1),
+    FowlCount_Median = round(median(waterFowl, na.rm = TRUE), 1),
+    FowlCount_Max = round(max(waterFowl, na.rm = TRUE), 1),
+
+    Turbidity_Mean = round(mean(turbidity, na.rm = TRUE), 1),
+    Turbidity_SD = round(sd(turbidity, na.rm = TRUE), 1),
+    Turbidity_Min = round(min(turbidity, na.rm = TRUE), 1),
+    Turbidity_Median = round(median(turbidity, na.rm = TRUE), 1),
+    Turbidity_Max = round(max(turbidity, na.rm = TRUE), 1)
+  ) %>%
+  pivot_longer(everything(), 
+               names_to = c("Variable", ".value"), 
+               names_pattern = "(.*)_(.*)")
+
+# Display the summary statistics table in a vertically aligned format
+kbl(summary_stats, col.names = c("Variable", "Mean", "SD", "Min", "Median", "Max")) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
+                full_width = F, 
+                position = "center", 
+                font_size = 10) %>%
+  row_spec(0, bold = TRUE)
 

### 2024-09-27 starter_folder-main/scripts/00-simulate_data.R

@@ -1,19 +1,51 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates Data
+# Author: Ziyuan Shen
+# Date: 22 September 2024 
+# Contact: ziyuan.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+# Define the date range (from mid-May to mid-September as per the dataset)
+start_date <- as.Date("2020-05-15") 
+end_date <- as.Date("2020-09-15")
+
+# Create a sequence of dates between start_date and end_date
+date_sequence <- seq.Date(start_date, end_date, by = "day")
+
+# Simulate random dates for 5 beaches (simulate for 500 observations in total)
+n <- 10  # Number of observations per beach
+beaches <- c("Beach1", "Beach2", "Beach3", "Beach4", "Beach5")
+random_dates <- sample(date_sequence, size = n * length(beaches), replace = TRUE)
+
+# Simulate data for the other variables
+simulated_data <- tibble(
+  id = 1:(n * length(beaches)),  # Create a unique ID for each observation
+  dataCollectionDate = random_dates,
+  beachName = rep(beaches, each = n),
+  windSpeed = round(runif(n * length(beaches), min = 0, max = 50), 1),  # wind speed in km/h
+  windDirection = sample(c("East", "North-East", "North", "North-West", "West", "South-West", "South", "South-East"), 
+                         size = n * length(beaches), replace = TRUE),
+  airTemp = round(runif(n * length(beaches), min = 10, max = 35), 1),  # air temperature in Celsius
+  rain = sample(c("Yes", "No"), size = n * length(beaches), replace = TRUE, prob = c(0.3, 0.7)),
+  rainAmount = ifelse(rain == "Yes", round(runif(n * length(beaches), min = 0.1, max = 30), 1), 0),  # rain in mm
[…]

### 2024-09-27 starter_folder-main/scripts/01-download_data.R

@@ -1,26 +1,32 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from Open Data Toronto
+# Author:  Ziyuan Shen
+# Date: 22 September 2024 
+# Contact: ziyuan.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NONE
+# Any other information needed? NONE
 
 
 #### Workspace setup ####
 library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
-
-#### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
 
 
+# get package
+package <- show_package("toronto-beaches-observations")
+package
+
+# get all resources for this package
+resources <- list_package_resources("toronto-beaches-observations")
+
+# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+
+# load the first datastore resource as a sample
+data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+data
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+write_csv(data, "data/raw_data/raw_data.csv") 
 
-         

### 2024-09-27 starter_folder-main/scripts/02-data_cleaning.R

@@ -1,44 +1,49 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the Toronto beaches data into an analysis dataset
+# Author: Ziyuan Shen
+# Date: 22 September 2024 
+# Contact: ziyuan.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NONE
+# Any other information needed? NONE
 
-#### Workspace setup ####
+
+####Step 1: Workspace setup ####
 library(tidyverse)
 
-#### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+raw_data <- read_csv("data/raw_data/raw_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+#### Step 2: Select the eight key variables ####
+key_variables <- raw_data %>%
+  select(dataCollectionDate, beachName, windSpeed, airTemp, rain, waterTemp, waterFowl, turbidity)
 
[…]

### 2024-09-27 starter_folder-main/scripts/03-test_data.R

@@ -1,15 +1,45 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Sanity check of the data
+# Author: Ziyuan Shen
+# Date: 22 September 2024 
+# Contact: ziyuan.shen@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NONE
+# Any other information needed? NONE
+
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Test data ####
+data <- read_csv("data/raw_data/simulated.csv")
+
+# Test NAs
+any(is.na(cleaned_data))
+
+# Remove NAs
+cleaned_data <- na.omit(data)
+
+
+# Test for NAs
+all(is.na(cleaned_data)) # If there is missing values in our raw dataset
+
+
+## function for testing the negative numbers
+check_non_negative <- function(data) {
+  negative_test <- data %>%
+    summarise_all(~ min(.)) %>%
+    gather(key = "column", value = "min_value") %>%
+    filter(min_value < 0)
+  if (nrow(negative_test) == 0) {
+    print("All columns have non-negative values.")
+  } else {
+    print("Negative values found in the following columns:") ##print when there's negative numbers
+    print(negative_test)
+  }
+}
+
+## check raw data
+check_non_negative(data)

### 2024-09-27 starter_folder-main/paper/paper.qmd
update base on peer reviews

@@
 title: "Environmental Factors Affecting Water Quality at Toronto Beaches"
-subtitle: "Investigation into Seasonal Trends(2010-2023)"
+subtitle: "In each summer period (2010-2023)"
 author: 
@@
 thanks: "Code and data are available at: https://github.com/Serena-SHEN1011/Toronto-Beach-Data"
-date: 23 September 2024
+date: 26 September 2024
 date-format: long
-abstract: "This paper investigates the environmental factors that influence water quality at Toronto beaches using summer data collected from 2010 to 2023.
-The research shows key variables such as rainfall, water temperature and waterfowl populations that significantly impact water quality safety warnings and public health.
-The study found that rainfall and water temperature fluctuations were strongly associated with beach warnings, emphasizing the necessity for improved safety measures.
-This research will help improve management strategies for Toronto's beaches and ensure a safer recreational experience for residents and visitors."
+abstract: "This study investigated the effects of various environmental factors, including rainfall, wind speed, air and water temperature, and the presence of waterfowl, on turbidity levels at Toronto beaches from 2010 to 2023.
+By analyzing the data, the study found that rainfall was the most important predictor of increased turbidity, while other factors such as air temperature and the presence of waterfowl showed a weaker correlation.
+The results of the study highlight the importance of monitoring rainfall events to maintain water quality and ensure public health safety at recreational beaches.
+This study contributes to a broader understanding of how environmental conditions affect water clarity and provides insights for policymakers to enhance beach management strategies."
 format: pdf
@@
 There are over 18,000 daily observations of beach conditions in this dataset.
-In the following section, I chose eight key environmental variables to examine in depth: date, beach name, wind speed, air temperature, water temperature, and waterfowl populations.
-By using these variables, we explore how environmental factors affect beach safety and availability.
-Specifically, we examined the relationship between wind speed and water quality, bird populations and water pollution risk, and how temperature variations affect the overall beach experience.
-This analysis will help to better plan and communicate the safety and attractiveness of Toronto's beaches.
-Structure is inspired mainly by @rohan.
-
-\newpage
+In the following section, I chose eight key environmental variables to examine in depth: date, beach name, wind speed, air temperature, water temperature, waterfowl populations and turbidity.
+The results of the study indicate that while rainfall is the main cause of increased turbidity, factors such as water temperature and the presence of waterfowl also play a role, albeit to a lesser extent.
+By identifying the conditions that contribute to poorer water quality, this study provides valuable insights for policymakers and beach management organizations, who can use these findings to develop more targeted interventions and recommendations.
+The methods used, the results obtained and their implications for public health and environmental management are described in detail in subsequent sections.
+And the following structure was inspired mainly by @rohan.
 
@@
 This decision was made to avoid any potential biases or inaccuracies that might result from incomplete data, as missing values could skew the results of the statistical analyses.
-By removing these rows, the dataset remained consistent and could be used for further analysis without introducing unnecessary complexity in dealing with missing data.
+Also, I removed several outliers.
+Some of the outliers such as the water temperature was up to 220 degree centigrade, which is a very unusual value.
+So I set reasonable range values for water temperature, air temperature, and wind speed, and removed any unreasonable values.
 
@@
 
-In addition, I have carefully selected nine key variables that were deemed most valuable to this analysis.
-These variables included date (the specific date on which observations were recorded) and beach name (identifying the different Toronto beaches on which environmental data were collected).
+Thus, I selected eight key variables that were deemed most valuable to this analysis.
+These variables included date and beach name.
 Wind speed and air temperature were listed as key weather-related variables, as these factors have a significant impact on beach conditions and overall water quality.
@@
 ```
-
 ```{r}
@@
 #| message: false
-#| label: tbl-sum5
-#| fig-cap: ""
+#| label: tbl-sum2
+#| fig-cap: "Summary statistics table"
 
@@
 ```
+@tbl-sum2 summaries the numerical data.
+The average wind speed was 13.3km/h with a maximum of 50 km/h, which stirs up sediments and may increase water turbidity.
+The average air temperature was 19.9°C, which directly affects the comfort of beachgoers and water quality, since warm air leads to higher water temperatures and increased microbial activity.
+Water temperatures averaged 15.6°C, reaching a maximum of 27°C during the summer months.
+this is often associated with bacterial growth, leading to health warnings.
+The number of waterfowl is another important factor, with an average of 32.3 waterfowl, but sometimes the number spikes to 450.
+The presence of waterfowl can bring faecal pollution into the water which increases the risk of bacterial contamination and water quality warnings.
 
-## Data Visualization {#sec3-data} 
-  Air temperature is a key indicator of beach conditions.
-@fig-his1 visualizes the distribution of air temperature at Toronto beaches and the count of observations for different temperature ranges.
-The x-axis represents air temperature in degrees Celsius (°C), while the y-axis represents the count of observations.
-The distribution is approximately normal, centered around 20°C, indicating that most observations were recorded in this temperature range.
-There are fewer observations in the lower and higher temperature ranges, with temperatures ranging from around 5°C to nearly 35°C.
-  The bar heights reflect the number of observations within each 2°C bin, showing that the majority of temperatures fall between 15°C and 25°C, typical of summer conditions.
-The steelblue color used for the bars provides a clear contrast against the black outlines, making the data easily distinguishable.
-This histogram effectively summarizes the air temperature data, revealing that Toronto beaches experience mild to warm temperatures during the observation period.
+\newpage
+
+# Result-Data Visualization {#third-result}
+## Turbidity across different beach {#third1-result} 
+@fig-box1 provides a clear visualization of the distribution of turbidity across various beach locations.
+Bluffer's Beach Park, Kew Balmy Beach, and Cherry Beach appear to have the highest overall turbidity levels, with numerous outliers reaching above 50 NTU and some exceeding 100 NTU.
+Hanlan’s Point Beach, Ward’s Island Beach, Gibraltar Point Beach, and Centre Island Beach exhibit much lower turbidity levels overall.
+Other beaches are showing more variability than low-turbidity beaches but less than the higher-turbidity ones.
 ```{r}
@@
 #| message: false
-#| label: fig-his1
-#| fig-cap: "Air Temperature Distribution"
+#| label: fig-box1
+#| fig-cap: "Distribution of Turbidity by Beach"
 
-ggplot(cleaned_data, aes(x = airTemp)) +
-  geom_histogram(binwidth = 2, fill = "steelblue", color = "black") +
-  labs(x = "Air Temperature (°C)", y = "Count") +
-  theme_minimal()
+# Boxplot of turbidity by beach
+ggplot(key_variables, aes(x = reorder(beachName, turbidity, FUN = median), y = turbidity)) +
+  geom_boxplot(fill = "skyblue", color = "black", outlier.color = "red", outlier.shape = 16) +
+  labs(x = "Beach Name", y = "Turbidity (NTU)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
 
@@
 
-  Wind speed affects both the comfort of beachgoers and water quality.
-@fig-bar1 illustrates the average wind speed at different Toronto beaches.
-The x-axis represents the average wind speed in kilometers per hour (km/h), while the y-axis lists the names of the beaches.
-The chart is horizontally oriented for clarity, with the wind speed values increasing from left to right.The lightblue bars represent the average wind speed at each beach, and the bars are arranged in descending order of wind speed.
-The chart highlights the variability in wind conditions across different beaches, suggesting that some beaches are more exposed to wind, potentially affecting people's experiences and water conditions.
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-#| label: fig-bar1
-#| fig-cap: "Average Wind Speed by Beach"
-
-ggplot(cleaned_data %>% group_by(beachName) %>% summarize(avg_wind = mean(windSpeed, na.rm = TRUE)), 
-       aes(x = reorder(beachName, avg_wind), y = avg_wind)) +
-  geom_bar(stat = "identity", fill = "lightblue") +
-  labs(x = "Beach", y = "Average Wind Speed (km/h)") +
-  coord_flip() + 
-  theme_minimal()
-
-```
-
-  Water temperature varies between beaches, which may influence swimmers' comfort and number of waterfowl in the water.
-@fig-bar2 shows the average water temperature at different Toronto beaches, with the x-axis representing the average water temperature in degrees Celsius (°C) and the y-axis listing the beach names.
-The chart is color-coded to distinguish each beach, and the bars are arranged in descending order of average water temperature.
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-#| label: fig-bar2
-#| fig-cap: "Average water temperature by beach"
-#| fig-subcap: ["Across different Toronto beaches"]
-
-ggplot(cleaned_data %>% group_by(beachName) %>% summarize(avg_water_temp = mean(waterTemp, na.rm = TRUE)), 
-       aes(x = reorder(beachName, avg_water_temp), y = avg_water_temp, fill = beachName)) +
-  geom_col() +
-  labs(x = "Beach", y = "Average Water Temperature (°C)") +
-  coord_flip() +
-  theme_minimal()  
-```
-
-  @fig-bar3 shows the total waterfowl count on rainy versus non-rainy days.
-The x-axis represents the presence or absence of rain, with three categories: "N/A" (unknown rain status), "No" (non-rainy days), and "Yes" (rainy days).
-The y-axis represents the total number of waterfowl observed.
-The chart indicates a noticeable difference in waterfowl presence based on weather conditions, with more waterfowl observed on non-rainy days compared to rainy ones.
-We can see that on non-rainy days (labeled "No"), the highest number of waterfowl were observed, with a count exceeding 60,000.
-On rainy days (labeled "Yes"), the total waterfowl count is significantly lower, around 40,000.
-The "N/A" category, representing days with unknown rain status, shows very few waterfowl, with a total count close to zero.
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-#| label: fig-bar3
-#| fig-cap: "Total water fowl count on rainy vs non-rainy days"
-#| results: hide  # This hides the console output but keeps the plot visible
-
-# 2.# Bar plot: Total water fowl count on rainy vs non-rainy days
-cleaned_data <- cleaned_data %>%
-  mutate(rain = ifelse(is.na(rain), "Unknown", rain))  # Convert NA values to "Unknown"
-
-# Checking the unique values of rain column
-unique(cleaned_data$rain)
-
-# Bar plot: Total water fowl count on rainy vs non-rainy days
-fowl_rain_plot <- cleaned_data %>%
-  group_by(rain) %>%
-  summarize(total_water_fowl = sum(waterFowl, na.rm = TRUE)) %>%
-  ggplot(aes(x = rain, y = total_water_fowl, fill = rain)) +
-  geom_col() +
-  labs(title = "Total Water Fowl Count on Rainy vs. Non-Rainy Days",
-       subtitle = "Comparing the total number of water fowls observed on rainy vs. non-rainy days",
-       x = "Rain (Yes/No)", y = "Total Water Fowl Count") +
-  scale_fill_manual(values = c("lightblue", "lightgreen", "gray")) +
-  theme_minimal() +
-  theme(legend.position = "none")
-
-# Print plot and caption
-print(fowl_rain_plot)
-
-```
-  @fig-scatter1 illustrates the relationship between wind speed (x-axis, measured in km/h) and turbidity (y-axis, measured in NTU).
-Each point represents an observation, with wind speed along the horizontal axis and turbidity along the vertical axis.
-This scatter plot shows that there is no strong linear relationship between wind speed and turbidity.
-However, lower wind speeds tend to be associated with a wide range of turbidity levels, whereas higher wind speeds appear to be less common and do not show higher turbidity levels.
-Most data points are clustered in areas with lower wind speeds (below 20 km/h) and turbidity levels below 50 NTU.
-Some outliers show higher turbidity levels above 150 NTU, especially at lower wind speeds.
-At wind speeds above 20 km/h, the data points become scattered, with very few data points over 40 km/h.
+## Relationship between Air Temperature and Turbidity {#third2-result} 
+@fig-scatter1 shows a flat linear regression line, indicating that turbidity remains relatively unaffected by changes in air temperature.
 ```{r}
@@
 #| label: fig-scatter1
-#| fig-cap: "Wind speed vs turbidity"
-#| fig-subcap: ["Analyzing the relationship between wind speed and water turbidity"]
+#| fig-cap: "Relationship between Air Temperature and Turbidity"
 
-# Scatter plot: Wind speed vs turbidity
-wind_turbidity_plot <- ggplot(cleaned_data, aes(x = windSpeed, y = turbidity)) +
-  geom_point(alpha = 0.6, color = "orange") +
-  labs(x = "Wind Speed (km/h)", y = "Turbidity (NTU)") +
+# Scatter plot of air temperature vs turbidity
+ggplot(key_variables, aes(x = airTemp, y = turbidity)) +
+  geom_point(alpha = 0.6) +
+  geom_smooth(method = "lm", col = "green") +
+  labs(x = "Air Temperature (°C)", y = "Turbidity (NTU)") +
   theme_minimal()
 
-# Print plot and caption
-print(wind_turbidity_plot)
 ```
 
-  Rainfall significantly impacts water turbidity, as rain can introduce pollutants and sediments into the water.
-@fig-box1 shows the distribution of turbidity levels (measured in NTU) on rainy versus non-rainy days.
-The x-axis represents whether rain was present ("Yes"), absent ("No"), or unknown ("N/A"), while the y-axis represents turbidity levels.
-The box plot suggests that most observations, whether rainy or not, tend to have lower turbidity levels.
-Non-rainy days exhibit a slightly broader spread of turbidity levels, but the majority of observations have relatively low turbidity (close to 0 NTU), with several outliers reaching beyond 200 NTU.
-Turbidity levels were also mostly low in Rainy (Yes) and N/A(Unknown), but there were outliers over 150 NTU.
+## Relationship between Water Temperature and Turbidity {#third3-result} 
+Relationship between Water Temperature and Turbidity shows the similar trend to @fig-scatter1.
+Both @fig-scatter1 and @fig-scatter2 shows that air temperature and water temperature are not a strong predictor of turbidity.
 ```{r}
@@
 #| message: false
-#| label: fig-box1
+#| label: fig-scatter2
+#| fig-cap: "Relationship between Water Temperature and Turbidity"
+
+# Scatter plot of water temperature vs turbidity
+ggplot(key_variables, aes(x = waterTemp, y = turbidity)) +
+  geom_point(alpha = 0.6) +
+  geom_smooth(method = "lm", col = "purple") +
+  labs(x = "water Temperature (°C)", y = "Turbidity (NTU)") +
+  theme_minimal()
+
+```
+
+## Relationship between Wind Speed and Turbidity{#third4-result} 
+@fig-scatter3 illustrates the relationship between wind speed (x-axis, measured in km/h) and turbidity (y-axis, measured in NTU).
+This scatter plot shows that there is no strong linear relationship between wind speed and turbidity.
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: fig-scatter3
+#| fig-cap: "Relationship between Wind Speed and Turbidity"
+
+# Scatter plot of wind speed vs turbidity
+ggplot(key_variables, aes(x = windSpeed, y = turbidity)) +
+  geom_point(alpha = 0.6) +
+  geom_smooth(method = "lm", col = "red") +
+  labs(x = "Wind Speed (km/h)", y = "Turbidity (NTU)") +
+  theme_minimal()
+```
+
+## Relationship between Waterfowl Count and Turbidity{#third5-result} 
+@fig-scatter4 is a scatter plot demonstrates the relationship between Waterfowl Count and Turbidity (measured in NTU).
+There is a slight upward trend, indicated by the blue linear regression line.
+This suggests that as the number of waterfowl increases, the turbidity also tends to rise, although the increase is relatively mild.
+Most of the data points are clustered around lower turbidity values (0-50 NTU), especially when waterfowl counts are below 100.
+This suggests that in many cases the presence of birds does not immediately lead to significant turbidity in the water column.
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: fig-scatter4
+#| fig-cap: "Relationship between Waterfowl Count and Turbidity"
+
+# Scatter plot of waterfowl count vs turbidity
+ggplot(key_variables, aes(x = waterFowl, y = turbidity)) +
+  geom_point(alpha = 0.6) +
+  geom_smooth(method = "lm", col = "blue") +
+  labs(x = "Waterfowl Count", y = "Turbidity (NTU)") +
+  theme_minimal()
+```
+Since there are many small values and some large outliers, I try to consider using a logarithmic scale for both waterfowl counts and turbidity to spread out the data and make trends more apparent.
+As shown in @fig-scatter5, it provides a better visualization of the data and suggests that while bird activity has some impact on water turbidity, it is not the only determinant.
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: fig-scatter5
+#| fig-cap: "Relationship between Waterfowl Count and Turbidity"
+
+# Scatter plot of waterfowl count vs turbidity with log scale
+ggplot(key_variables, aes(x = waterFowl, y = turbidity)) +
+  geom_point(alpha = 0.6) +
+  geom_smooth(method = "lm", col = "blue") +
+  scale_x_log10() +  # Log scale for waterfowl count
+  scale_y_log10() +  # Log scale for turbidity
+  labs(x = "Waterfowl Count (Log Scale)", y = "Turbidity (NTU, Log Scale)") +
+  theme_minimal()
+
+```
+
+## Turbidity Levels on Rainy vs. Non-Rainy Day{#third6-result} 
+@fig-box2 shows the distribution of turbidity levels (measured in NTU) on rainy versus non-rainy days.
+On rainy days, the change in turbidity increases relatively significantly.
+The occurrence of multiple outliers above 50 NTU indicates that rainfall is largely responsible for the increase in turbidity.
+The larger interquartile range (IQR) indicates that turbidity values are more scattered on rainy days compared to non-rainy days.
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: fig-box2
 #| fig-cap: "Turbidity Levels on Rainy vs. Non-Rainy Days"
@@
 
-# Discussion{#third-dis}
-## Data Collection and Limitations
+# Discussion{#fourth-dis}
+## Impacts of beach location on water quality{#fourth1-dis}
+When analyzing the relationship between beach location and turbidity, Bluffer's Beach Park, Kew Balmy Beach and Cherry Beach were consistently higher compared to the other locations.
+These beaches not only had more outliers but also had larger interquartile ratios (IQRs), indicating large fluctuations in water quality over time.
+The high turbidity at these beaches may be due to their geographic location, proximity to sources of urban runoff, or local environmental conditions that increase the potential for pollution.
+
+## Impact of Rainfall on Water Quality{#fourth2-dis}
+The above observations suggest that rainfall is an important factor in influencing turbidity levels in Toronto's beach waters.
+The study found that on rainy days, turbidity levels increased significantly, with several outliers exceeding 100 NTU, suggesting that rainfall may have washed pollutants into the water and reduced clarity.
+This finding underscores the role of rainfall in carrying sediment, pollutants, and other debris into water, leading to potential public health risks.
+The large interquartile range on rainy days suggests that water quality can be highly variable, further stressing the need for a robust water quality monitoring system following heavy rainfall.
+
+## The Role of Waterfowl in Turbidity{#fourth3-dis}
+When examining the relationship between waterfowl and turbidity levels, the correlation between the two was weak, although a slight upward trend was observed.
+The data showed that in most cases, the presence of waterfowl did not result in an immediate spike in turbidity levels, suggesting that the impact of waterfowl is more cumulative or dependent on other environmental factors such as rainfall.
+However, the presence of large numbers of waterfowl does pose a pollution risk as they can carry faecal contaminants into the water and elevate bacterial levels.
+Therefore, I think more data and information on the impact of waterfowl can be sought for further study.
+
+## Wind speed and its effect on turbidity{#fourth4-dis}
+It is often assumed that wind speed affects water quality by stirring up sediments at the bottom of the water.
+However, research results have shown no strong relationship between wind speed and turbidity.
+While lower wind speeds were associated with a wider range of turbidity, higher wind speeds did not necessarily result in higher turbidity.
+This suggests that while wind is important, it may not be the primary driver of water clarity at these beaches.
+
+## Temperature and Turbidity{#fourth5-dis}
+It was found that both air and water temperatures had little effect on turbidity.
+The flat regression line in the temperature turbidity analysis showed that temperature fluctuations by themselves do not significantly affect changes in water clarity.
+This was a little surprising to me because higher temperatures tend to be associated with increased microbial activity, and increased microbial activity can affect water quality.
+
+## Data Limitations and future improvement{#fourth6-dis}
 Observations in the dataset are obtained by city staff through a combination of both manual estimates and instrumental measurements.
-While the use of automatic tools such as thermometers and turbidimeters can provide more objective and consistent data, manual observations such as waterfowl counts and qualitative assessments of water clarity introduce an element of subjectivity.
-In addition, removing outliers and missing data entries during the clean-up process addresses errors that can occur during data collection, especially extreme or unreasonable values (e.g., water temperatures above 200°C).
+Manual data collection introduces potential bias and inconsistency because human judgement can vary and is more prone to observational errors.
+In addition, while some unusual outliers and missing data improves the completeness of the dataset, these filters may inadvertently omit important rare events or distort trends.
+For example, extreme weather events or unusual environmental conditions (e.g., large bacterial outbreaks) may have been filtered out during the data cleaning process.
 
-In summary, we measured real-world temperature, speed of wind, rainfall, turbidity, and waterfowl activity using a variety of instruments and estimation techniques, resulting in quantifiable data entries in the dataset.
-These measurements form the basis for analysing environmental conditions at Toronto Beach, and transform natural phenomena into structured data that can be analysed to reveal trends, patterns and relationships between different variables.
+Another limiting factor is the lack of detailed water quality measurements other than turbidity.
+While turbidity is a useful proxy for water clarity, it does not provide a complete picture of water contamination or health risks.
+Bacterial contamination may occur in relatively clear water, and the lack of a strong relationship between water temperature and turbidity suggests that turbidity alone may not adequately reflect microbial activity.
+So I expect that future research will provide a more complete understanding of public health risks by incorporating additional water quality indicators, such as bacterial counts or chemical contaminants.
 
@@
 #| message: false
-#| label: tbl-sum2
+#| label: tbl-sum3
 #| fig-cap: "Description of variables in the dataset"

### 2024-09-27 starter_folder-main/paper/paper.qmd
update subtitle

@@
 
-\newpage
-
-# Result-Data Visualization {#third-result}
-## Turbidity across different beach {#third1-result} 
+##Data Visualization {#sec3-data} 
+### Turbidity across different beach 
 @fig-box1 provides a clear visualization of the distribution of turbidity across various beach locations.
@@
 
-## Relationship between Air Temperature and Turbidity {#third2-result} 
+### Relationship between Air Temperature and Turbidity 
 @fig-scatter1 shows a flat linear regression line, indicating that turbidity remains relatively unaffected by changes in air temperature.
@@
 
-## Relationship between Water Temperature and Turbidity {#third3-result} 
+### Relationship between Water Temperature and Turbidity 
 Relationship between Water Temperature and Turbidity shows the similar trend to @fig-scatter1.
@@
 
-## Relationship between Wind Speed and Turbidity{#third4-result} 
+### Relationship between Wind Speed and Turbidity
 @fig-scatter3 illustrates the relationship between wind speed (x-axis, measured in km/h) and turbidity (y-axis, measured in NTU).
@@
 
-## Relationship between Waterfowl Count and Turbidity{#third5-result} 
+### Relationship between Waterfowl Count and Turbidity
 @fig-scatter4 is a scatter plot demonstrates the relationship between Waterfowl Count and Turbidity (measured in NTU).
@@
 
-## Turbidity Levels on Rainy vs. Non-Rainy Day{#third6-result} 
+### Turbidity Levels on Rainy vs. Non-Rainy Day
 @fig-box2 shows the distribution of turbidity levels (measured in NTU) on rainy versus non-rainy days.
@@
 
-# Discussion{#fourth-dis}
-## Impacts of beach location on water quality{#fourth1-dis}
+# Discussion{#third-dis}
+## Impacts of beach location on water quality{#third1-dis}
 When analyzing the relationship between beach location and turbidity, Bluffer's Beach Park, Kew Balmy Beach and Cherry Beach were consistently higher compared to the other locations.
@@
 
-## Impact of Rainfall on Water Quality{#fourth2-dis}
+## Impact of Rainfall on Water Quality{#third2-dis}
 The above observations suggest that rainfall is an important factor in influencing turbidity levels in Toronto's beach waters.
@@
 
-## The Role of Waterfowl in Turbidity{#fourth3-dis}
+## The Role of Waterfowl in Turbidity{#third3-dis}
 When examining the relationship between waterfowl and turbidity levels, the correlation between the two was weak, although a slight upward trend was observed.
@@
 
-## Wind speed and its effect on turbidity{#fourth4-dis}
+## Wind speed and its effect on turbidity{#third4-dis}
 It is often assumed that wind speed affects water quality by stirring up sediments at the bottom of the water.
@@
 
-## Temperature and Turbidity{#fourth5-dis}
+## Temperature and Turbidity{#third5-dis}
 It was found that both air and water temperatures had little effect on turbidity.
@@
 
-## Data Limitations and future improvement{#fourth6-dis}
+## Data Limitations and future improvement{#third6-dis}
 Observations in the dataset are obtained by city staff through a combination of both manual estimates and instrumental measurements.

### 2024-09-27 paper/paper.qmd
update paper

--- 2024-09-27 starter_folder-main/paper/paper.qmd
@@
 library(tibble)
+library(dplyr)
+library(knitr)
 ```
@@
 cleaned_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+# choose 8 variables
+key_variables <- cleaned_data %>%
+  select(dataCollectionDate, beachName, windSpeed, airTemp, rain, waterTemp, waterFowl, turbidity)
+
 ```
@@
 The analysis focuses on the key variables affecting water turbidity, which is a proxy for water pollution risk.
-All variables are explained and described in (Section A.[1](#appendix-a.1)) In my this report, I will focuses on 8 key variables: date, beach name, wind's speed, air temperature, wether rain or not, water temperature, waterfowl, water turbidity.
+All key variables are explained and described in (Section A.[1](#appendix-a.1)) In my this report, I will focuses on 8 key variables: date, beach name, wind's speed, air temperature, wether rain or not, water temperature, waterfowl, water turbidity.
 
@@
 
-## Cleaned Data {#sec2-data} 
-During the data cleaning process, I took several important steps to ensure that the dataset was suitable for analysis.
-First, all rows containing missing values (NA) were removed.
-This decision was made to avoid any potential biases or inaccuracies that might result from incomplete data, as missing values could skew the results of the statistical analyses.
-Also, I removed several outliers.
-Some of the outliers such as the water temperature was up to 220 degree centigrade, which is a very unusual value.
-So I set reasonable range values for water temperature, air temperature, and wind speed, and removed any unreasonable values.
+I cleaned and prepared a subset of the data as shown in (Section A.[2](#appendix-a.2)).
+(Section A.[2](#appendix-a.2)) described the process of my data cleaning.
+I showed the first and last 5 rows of the key variables in @tbl-sum2.
+Also, I summarize the numerical data in @tbl-sum3.
 
-At the same time, several variables were removed because they were either redundant or irrelevant to the focus of this analysis.
-Wind direction was excluded because it has less impact on water conditions than wind speed.
-Similarly, rainfall was replaced by a simplified binary rainfall variable that indicates whether or not it rained on a given day.
-This binary approach was chosen to simplify the analysis without compromising the central idea that rainfall affects water quality.
-The role of waves was removed.
-This is because, although wave intensity affects beach conditions, its overall effect is not as important to the current discussion of weather, waterfowl, and water quality.
-
-Thus, I selected eight key variables that were deemed most valuable to this analysis.
-These variables included date and beach name.
-Wind speed and air temperature were listed as key weather-related variables, as these factors have a significant impact on beach conditions and overall water quality.
-Water temperature was also considered critical because of its impact on both the beachgoer experience and biological activity in the water.
-Waterfowl population and turbidity were chosen because they are directly related to water quality.
-Waterfowl can cause pollution, while turbidity measures water clarity and is often used as an indicator of pollution.
-Finally, the presence or non-presence of rain was included to capture binary rainfall scenarios, as rainfall can affect water clarity and carry pollutants into the beach environment.
-
-I cleaned and prepared a subset of the data as shown in @tbl-sum1 below. I showed the first and last 5 rows of the key variables in this table.
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-#| label: tbl-sum1
-#| fig-cap: "Displaying the first and last 5 rows of seven key variables"
-
-# choose 8 variables
-key_variables <- cleaned_data %>%
-  select(dataCollectionDate, beachName, windSpeed, airTemp, rain, waterTemp, waterFowl, turbidity)
-
-# Displaying the first and last 5 rows of key variables
-sample_data <- bind_rows(head(key_variables, 5), tail(key_variables, 5))
-
-# Generate table
-kbl(sample_data, 
-    col.names = c("Date", "Beach", "Wind (km/h)", "Air Temp (°C)", "Rain", "Water Temp (°C)", "Fowl Count", "Turbidity (NTU)"),
-    ) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                position = "center",
-                font_size = 9) %>%
-  column_spec(1, bold = TRUE, width = "6em") %>%  
-  column_spec(2, width = "8em") %>%  
-  column_spec(3:4, width = "5em") %>%  
-  column_spec(5, width = "2em") %>%  
-  column_spec(6, width = "6em") %>%
-  column_spec(7:8, width = "3em")
-
-```
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-#| label: tbl-sum2
-#| fig-cap: "Summary statistics table"
-
-# Calculate summary statistics for key variables
-summary_stats <- key_variables %>%
-  summarise(
-    WindSpeed_Mean = round(mean(windSpeed, na.rm = TRUE), 1),
-    WindSpeed_SD = round(sd(windSpeed, na.rm = TRUE), 1),
-    WindSpeed_Min = round(min(windSpeed, na.rm = TRUE), 1),
-    WindSpeed_Median = round(median(windSpeed, na.rm = TRUE), 1),
-    WindSpeed_Max = round(max(windSpeed, na.rm = TRUE), 1),
-
-    AirTemp_Mean = round(mean(airTemp, na.rm = TRUE), 1),
-    AirTemp_SD = round(sd(airTemp, na.rm = TRUE), 1),
-    AirTemp_Min = round(min(airTemp, na.rm = TRUE), 1),
-    AirTemp_Median = round(median(airTemp, na.rm = TRUE), 1),
-    AirTemp_Max = round(max(airTemp, na.rm = TRUE), 1),
-
-    WaterTemp_Mean = round(mean(waterTemp, na.rm = TRUE), 1),
-    WaterTemp_SD = round(sd(waterTemp, na.rm = TRUE), 1),
-    WaterTemp_Min = round(min(waterTemp, na.rm = TRUE), 1),
-    WaterTemp_Median = round(median(waterTemp, na.rm = TRUE), 1),
-    WaterTemp_Max = round(max(waterTemp, na.rm = TRUE), 1),
-
-    FowlCount_Mean = round(mean(waterFowl, na.rm = TRUE), 1),
-    FowlCount_SD = round(sd(waterFowl, na.rm = TRUE), 1),
-    FowlCount_Min = round(min(waterFowl, na.rm = TRUE), 1),
-    FowlCount_Median = round(median(waterFowl, na.rm = TRUE), 1),
-    FowlCount_Max = round(max(waterFowl, na.rm = TRUE), 1),
-
-    Turbidity_Mean = round(mean(turbidity, na.rm = TRUE), 1),
-    Turbidity_SD = round(sd(turbidity, na.rm = TRUE), 1),
-    Turbidity_Min = round(min(turbidity, na.rm = TRUE), 1),
-    Turbidity_Median = round(median(turbidity, na.rm = TRUE), 1),
-    Turbidity_Max = round(max(turbidity, na.rm = TRUE), 1)
-  ) %>%
-  pivot_longer(everything(), 
-               names_to = c("Variable", ".value"), 
-               names_pattern = "(.*)_(.*)")
-
-# Display the summary statistics table in a vertically aligned format
-kbl(summary_stats, col.names = c("Variable", "Mean", "SD", "Min", "Median", "Max")) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
-                full_width = F, 
-                position = "center", 
-                font_size = 10) %>%
-  row_spec(0, bold = TRUE)
-
-```
-@tbl-sum2 summaries the numerical data.
-The average wind speed was 13.3km/h with a maximum of 50 km/h, which stirs up sediments and may increase water turbidity.
-The average air temperature was 19.9°C, which directly affects the comfort of beachgoers and water quality, since warm air leads to higher water temperatures and increased microbial activity.
-Water temperatures averaged 15.6°C, reaching a maximum of 27°C during the summer months.
-this is often associated with bacterial growth, leading to health warnings.
-The number of waterfowl is another important factor, with an average of 32.3 waterfowl, but sometimes the number spikes to 450.
-The presence of waterfowl can bring faecal pollution into the water which increases the risk of bacterial contamination and water quality warnings.
-
-##Data Visualization {#sec3-data} 
+## Data Visualization{#sec2-data}
 ### Turbidity across different beach 
-@fig-box1 provides a clear visualization of the distribution of turbidity across various beach locations.
+@fig-box1 provides a visualization of the distribution of turbidity across various beach locations.
+Since this diagram looked squeezed, I used a logarithmic scale for to make trends more apparent @fig-box2.
+The new diagram looks to contrast very well.
 Bluffer's Beach Park, Kew Balmy Beach, and Cherry Beach appear to have the highest overall turbidity levels, with numerous outliers reaching above 50 NTU and some exceeding 100 NTU.
@@
 ggplot(key_variables, aes(x = reorder(beachName, turbidity, FUN = median), y = turbidity)) +
-  geom_boxplot(fill = "skyblue", color = "black", outlier.color = "red", outlier.shape = 16) +
+  geom_boxplot(fill = "cyan", color = "black", outlier.color = "red", outlier.shape = 16) +
   labs(x = "Beach Name", y = "Turbidity (NTU)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: fig-box2
+#| fig-cap: "Distribution of Turbidity by Beach (Log Scale)"
+
+# Boxplot of turbidity by beach with log scale
+ggplot(key_variables, aes(x = reorder(beachName, turbidity, FUN = median), y = turbidity)) + 
+  geom_boxplot(fill = "cyan", color = "black", outlier.color = "red", outlier.shape = 16) +
+  scale_y_log10() +  # Apply log scale transformation to y-axis
+  labs(x = "Beach Name", y = "Turbidity (NTU, Log Scale)") +
   theme_minimal() +
@@
 #| message: false
-#| label: fig-box2
+#| label: fig-box3
 #| fig-cap: "Turbidity Levels on Rainy vs. Non-Rainy Days"
@@
 #| message: false
-#| label: tbl-sum3
-#| fig-cap: "Description of variables in the dataset"
+#| label: tbl-sum1
+#| fig-cap: "Variable description"
 
 data <- data.frame(
-  column = c("_id", "dataCollectionDate", "beachName", "windSpeed", 
-             "windDirection", "airTemp", "rain", "rainAmount", "waterTemp", 
-             "waterFowl", "waveAction", "waterClarity", "turbidity"),
-
-  description = c("Unique row identifier for Open Data database", 
+  Column = c("dataCollectionDate", "beachName", "windSpeed", "airTemp", "rain", "waterTemp", "waterFowl", "turbidity"),
+  Description = c(
     "Date observations were collected",
@@
     "Wind speed measured in km/h",
-                  "Wind directions - East, North-East, North, North-West, West, South-West, South, South-East", 
     "Air temperature in degrees Celsius",
     "Presence of rain - Yes, No",
-                  "Amount of rainfall in mm in the last 24 hours", 
     "Water temperature in degrees Celsius",
     "Count of water fowl seen",
-                  "Observations on wave action - high, low, moderate, none", 
-                  "Visual state of water's clarity - free form text", 
-                  "Turbidity of water measured in Nephelometric Turbidity Units (NTUs)")
+    "Turbidity of water measured in Nephelometric Turbidity Units (NTUs)"
   )
-data$description <- str_wrap(data$description, width = 25)
+)
 
-kable(data, 
-      align = "l", 
-      col.names = c("Column", "Description"),
-      escape = FALSE) %>%
-  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
+data |>
+  kable(col.names = c("Column", "Description"), align = c("l", "l"), escape = FALSE, booktabs = TRUE) |>
+  kable_styling(full_width = F, position = "center", font_size = 10) |>
+  row_spec(0, bold = TRUE, color = "black", background = "lightgray") |> 
+  column_spec(1, bold = TRUE, width = "15em") |> 
+  column_spec(2, width = "30em") |> 
+  footnote(general = "Note: Description of variables used in the Toronto beach water quality dataset.", general_title = " ")
+
+```
+
+## Cleaned Data {#appendix-a.2}
+During the data cleaning process, I took several important steps to ensure that the dataset was suitable for analysis.
+First, all rows containing missing values (NA) were removed.
+This decision was made to avoid any potential biases or inaccuracies that might result from incomplete data, as missing values could skew the results of the statistical analyses.
+Also, I removed several outliers.
+Some of the outliers such as the water temperature was up to 220 degree centigrade, which is a very unusual value.
+So I set reasonable range values for water temperature, air temperature, and wind speed, and removed any unreasonable values.
+
+At the same time, several variables were removed because they were either redundant or irrelevant to the focus of this analysis.
+Wind direction was excluded because it has less impact on water conditions than wind speed.
+Similarly, rainfall was replaced by a simplified binary rainfall variable that indicates whether or not it rained on a given day.
+This binary approach was chosen to simplify the analysis without compromising the central idea that rainfall affects water quality.
+The role of waves was removed.
+This is because, although wave intensity affects beach conditions, its overall effect is not as important to the current discussion of weather, waterfowl, and water quality.
+
+Thus, I selected eight key variables that were deemed most valuable to this analysis.
+These variables included date and beach name.
+Wind speed and air temperature were listed as key weather-related variables, as these factors have a significant impact on beach conditions and overall water quality.
+Water temperature was also considered critical because of its impact on both the beachgoer experience and biological activity in the water.
+Waterfowl population and turbidity were chosen because they are directly related to water quality.
+Waterfowl can cause pollution, while turbidity measures water clarity and is often used as an indicator of pollution.
+Finally, the presence or non-presence of rain was included to capture binary rainfall scenarios, as rainfall can affect water clarity and carry pollutants into the beach environment.
+
+@tbl-sum2 summaries the numerical data.
+The average wind speed was 13.3km/h with a maximum of 50 km/h, which stirs up sediments and may increase water turbidity.
+The average air temperature was 19.9°C, which directly affects the comfort of beachgoers and water quality, since warm air leads to higher water temperatures and increased microbial activity.
+Water temperatures averaged 15.6°C, reaching a maximum of 27°C during the summer months.
+this is often associated with bacterial growth, leading to health warnings.
+The number of waterfowl is another important factor, with an average of 32.3 waterfowl, but sometimes the number spikes to 450.
+The presence of waterfowl can bring faecal pollution into the water which increases the risk of bacterial contamination and water quality warnings.
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: tbl-sum2
+#| fig-cap: "Displaying the first and last 5 rows of seven key variables"
+
+# Displaying the first and last 5 rows of key variables
+sample_data <- bind_rows(head(key_variables, 5), tail(key_variables, 5))
+
+# Create the formatted table using kable and kableExtra
+sample_data %>%
+  kable(
+    col.names = c("Date", "Beach", "Wind (km/h)", "Air Temp (°C)", "Rain", "Water Temp (°C)", "Fowl Count", "Turbidity (NTU)"), # Column headers
+    digits = 1,            # Number of decimal places for numerical columns
+    booktabs = TRUE,       # Enable booktabs for improved styling
+    align = c("l", "l", "c", "c", "c", "c", "r", "r"),  # Align columns (left, center, right)
+    format.args = list(big.mark = ",")  # Use commas as thousand separators for large numbers
+  ) %>%
+  kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"), # Bootstrap styling options
+    full_width = FALSE,   # Prevent the table from taking full width
+    position = "center",  # Center align the table
+    font_size = 8        # Set font size
+  ) %>%
+  row_spec(0, bold = TRUE, background = "lightgray", color = "black") %>%  # Style header row
+  column_spec(1, width = "6em") %>%    # Set width and style for Date column
+  column_spec(2, width = "8em") %>%                 # Set width for Beach column
+  column_spec(3:4, width = "4em") %>%               # Set width for Wind and Air Temp columns
+  column_spec(5, width = "3em") %>%                 # Set width for Rain column
+  column_spec(6, width = "5em") %>%                 # Set width for Water Temp column
+  column_spec(7:8, width = "5em")                   # Set width for Fowl Count and Turbidity columns
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: tbl-sum3
+#| fig-cap: "Summary numerical variables"
+
+# Calculate summary statistics for key variables
+summary_stats <- key_variables %>%
+  summarise(
+    WindSpeed_Mean = round(mean(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_SD = round(sd(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_Min = round(min(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_Median = round(median(windSpeed, na.rm = TRUE), 1),
+    WindSpeed_Max = round(max(windSpeed, na.rm = TRUE), 1),
+
+    AirTemp_Mean = round(mean(airTemp, na.rm = TRUE), 1),
+    AirTemp_SD = round(sd(airTemp, na.rm = TRUE), 1),
+    AirTemp_Min = round(min(airTemp, na.rm = TRUE), 1),
+    AirTemp_Median = round(median(airTemp, na.rm = TRUE), 1),
+    AirTemp_Max = round(max(airTemp, na.rm = TRUE), 1),
+
+    WaterTemp_Mean = round(mean(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_SD = round(sd(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_Min = round(min(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_Median = round(median(waterTemp, na.rm = TRUE), 1),
+    WaterTemp_Max = round(max(waterTemp, na.rm = TRUE), 1),
+
+    FowlCount_Mean = round(mean(waterFowl, na.rm = TRUE), 1),
+    FowlCount_SD = round(sd(waterFowl, na.rm = TRUE), 1),
+    FowlCount_Min = round(min(waterFowl, na.rm = TRUE), 1),
+    FowlCount_Median = round(median(waterFowl, na.rm = TRUE), 1),
+    FowlCount_Max = round(max(waterFowl, na.rm = TRUE), 1),
+
+    Turbidity_Mean = round(mean(turbidity, na.rm = TRUE), 1),
+    Turbidity_SD = round(sd(turbidity, na.rm = TRUE), 1),
+    Turbidity_Min = round(min(turbidity, na.rm = TRUE), 1),
+    Turbidity_Median = round(median(turbidity, na.rm = TRUE), 1),
+    Turbidity_Max = round(max(turbidity, na.rm = TRUE), 1)
+  ) %>%
+  pivot_longer(everything(), 
+               names_to = c("Variable", ".value"), 
+               names_pattern = "(.*)_(.*)")
+
+summary_stats %>%
+  kable(
+    col.names = c("Variable", "Mean", "SD", "Min", "Median", "Max"), 
+    booktabs = TRUE,  
+    align = c("l", "c", "c", "c", "c", "c"),   
+    format.args = list(big.mark = ",")  
+  ) %>%
+  kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"), 
     full_width = F,   
     position = "center",   
-                font_size = 10)
+    font_size = 10   
+  ) %>%
+  row_spec(0, bold = TRUE, background = "lightgray", color = "black") %>%  # Style header row
+  footnote(general = "Summary statistics for Wind Speed, Air Temperature, Water Temperature, Fowl Count, and Turbidity.")
 ```

### 2024-09-27 paper/paper.qmd
Beautification tables

@@
 #| label: tbl-sum1
-#| fig-cap: "Variable description"
+#| fig-cap: "Key Variable Description"
 
@@
 ```
-
+\newpage
 ```{r}
@@
   row_spec(0, bold = TRUE, background = "lightgray", color = "black") %>%  # Style header row
+  column_spec(1, bold = TRUE, width = "9em") %>%    # Set width and style for Variable
+  column_spec(2:6, width = "6em") %>%    # Set width for the rest column
   footnote(general = "Summary statistics for Wind Speed, Air Temperature, Water Temperature, Fowl Count, and Turbidity.")
+
 ```
